---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Mode estimation in R

<!-- badges: start -->

[![R-CMD-check](https://github.com/lhdjung/moder/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/lhdjung/moder/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The moder package determines single or multiple **modes** (most frequent values). By default, its functions check whether missing values make this impossible, and return `NA` in this case. It has no dependencies.

Mode functions fill a gap in measures of central tendency in R. `mean()` and `median()` are built into the standard library, but there is a lack of go-to functions for calculating the mode. Earlier solutions don't handle missing values properly. Use moder for this!

## Installation

You can install the development version of moder like so:

``` r
remotes::install_github("lhdjung/moder")
```

## Get started

```{r}
library(moder)
```

### Find the first mode with `mode_first()`

Everything is fine here:

```{r}
mode_first(c(7, 8, 8, 9, 9, 9))
```

But what if some values are missing? Think about what it means to look for the most frequent value in such a distribution. Maybe there are so many missings that it's impossible to tell which value is the most frequent one:

```{r}
mode_first(c(1, 1, 2, 2, 2, 2, NA, NA, NA, NA))
```

Ignore `NA`s using `na.rm = TRUE` if there is a strong rationale for it:

```{r}
mode_first(c(1, 1, 2, 2, 2, 2, NA, NA, NA, NA), na.rm = TRUE)
```

This distribution is different:

```{r}
mode_first(c(7, 7, 7, 7, 8, 8, NA))
```

Even if the `NA` stands in for `8`, there will only be three instances of `8` but four instances of `7`. The mode is `7`, independent of the true value behind `NA`.

### Find all modes with `mode_all()`

This function captures multiple modes:

```{r}
mode_all(c("a", "a", "b", "b", "c", "d", "e"))
```

If some values are missing but there would be multiple modes when ignoring `NA`s, `mode_all()` returns `NA`. That's because missings can easily create an imbalance between the equally-frequent known values:

```{r}
mode_all(c(1, 1, 2, 2, NA))
```

If `NA` masks either `1` or `2`, that number is the (single) mode. As before, if the mode depends on missing values, the function returns `NA`.

Yet `na.rm = TRUE` makes the function ignore this:

```{r}
mode_all(c(1, 1, 2, 2, NA), na.rm = TRUE)
```

### Find the single mode (or `NA`) with `mode_single()`

`mode_single()` is stricter than `mode_first()`: It returns `NA` if there are multiple modes. Otherwise, it works the same way.

```{r}
mode_single(c(3, 4, 4, 5, 5, 5))
mode_single(c("x", "x", "y", "y", "z"))
```

### Find possible modes

These minimal and maximal sets of modes are possible given the missing value:

```{r}
mode_possible_min(c("a", "a", "a", "b", "b", "c", NA))
mode_possible_max(c("a", "a", "a", "b", "b", "c", NA))
```
