---
title: "Modal counts and frequencies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modal counts and frequencies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
# Dev only: load moder from within moder
devtools::load_all(".")
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(moder)
```

In addition to finding a vector's mode(s), you might be interested in some metadata about them:

-   A vector's modal **count** is the number of its modes.

-   A vector's modal **frequency** is the number of times that a vector's mode appears in the vector.

This vignette briefly presents moder's tools for determining these quantities.

## Modal count

`mode_count()` computes the number of modes:

```{r}
mode_count(c(5, 5, 6))
mode_count(c(5, 5, 6, 6, 7))
```

Even with missing values, the number of modes is sometimes known. It can only be 1 here because even if the `NA` is secretly `"b"`, then `"b"` would appear twice, but `"a"` would appear three times:

```{r}
mode_count(c("a", "a", "a", "b", NA))
```

All this only works if the full set of modes can be determined. Below, `NA` could secretly be `5`, `6`, or any other value. If it's `5` or `6`, that value is the mode but the other value isn't, so there is only one mode. But if it's some other value, `5` and `6` are equally frequent, and both are modes. Since we lack this information, the number of modes is unknown.

All this only works if the full set of modes can be determined. Below, `NA` could secretly be `7`, `8`, or any other value. If it's `8`, there are two modes, `7` and `8`. Otherwise, `7` is the only mode. Since we lack this information, the number of modes is unknown.

```{r}
mode_count(c(7, 7, 7, 8, 8, NA))

```

Use `mode_count_range()` in such cases. It will determine the minimal and maximal number of modes:

```{r}
mode_count_range(c(7, 7, 7, 8, 8, NA))
```

Yet even this function can fail because some vectors don't have a clear minimum or maximum. Here, `NA` could be either `1` or `2` in disguise, which would make that number the only mode:

```{r}
mode_count_range(c(1, 1, 2, 2, NA))
```

## Modal frequency

`mode_frequency()` counts the instances of a vector's modes in the vector:

```{r}
mode_frequency(c(4, 4, 5))
mode_frequency(c(4, 4, 4, 5))
```

Missing values are an issue here, even if the mode is obvious. Each `NA` might be another instance of the mode, so the frequency is unknown:

```{r}
mode_frequency(c(1, 1, 1, 1, 2, NA))
```

With `mode_frequency_range()`, at least the minimal and maximal frequencies can be determined. There are four instances of `1` without counting the `NA`, and five with counting it:

```{r}
mode_frequency_range(c(1, 1, 1, 1, 2, NA))
```

Finally, a function that never returns `NA`!
