---
title: "Modal counts and frequencies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modal counts and frequencies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
# Dev only: load moder from within moder
devtools::load_all(".")
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(moder)
```

In addition to finding a vector's modes, you might be interested in some metadata about them:

-   A vector's modal **count** is the number of its modes.

-   A vector's modal **frequency** is the number of times that any single mode appears in the vector.

## Modal count

`mode_count()` computes the number of modes:

```{r}
mode_count(c(5, 5, 6))
mode_count(c(5, 5, 6, 6, 7))
```

Even with missing values, the number of modes is sometimes known. It can only be 1 here because even if the `NA` is secretly `"b"`, then `"b"` would appear twice, but `"a"` would appear three times:

```{r}
mode_count(c("a", "a", "a", "b", NA))
```

All of this only works if the full set of modes can be determined. Below, `NA` could secretly be `7`, `8`, or any other value. If it's `8`, both numbers are equally frequent. Otherwise, `7` is the only mode. Since we lack this information, the number of modes is unknown.

```{r}
mode_count(c(7, 7, 7, 8, 8, NA))

```

Use `mode_count_range()` in such cases. It will determine the minimal and maximal number of modes, never returning `NA`. (For more on `mode_count_range()`, see `vignette("missings")`, section *Maximal number of unique values*.)

```{r}
mode_count_range(c(7, 7, 7, 8, 8, NA))
```

## Modal frequency

`mode_frequency()` counts the instances of a vector's modes in the vector:

```{r}
mode_frequency(c(4, 4, 5))
mode_frequency(c(4, 4, 4, 5))
```

Missing values are an issue here, even if the mode is obvious. Each `NA` might be another instance of the mode, so the frequency is unknown:

```{r}
mode_frequency(c(1, 1, 1, 1, 2, NA, NA))
```

With `mode_frequency_range()`, at least the minimal and maximal frequencies can be determined. It never returns `NA`. The minimum frequency supposes that no `NA`s represent the mode; the maximum frequency supposes that all of them do. In this way, there are four instances of `1` without counting the `NA`s, and six with counting them:

```{r}
mode_frequency_range(c(1, 1, 1, 1, 2, NA, NA))
```

### Trivial modes

Related to frequencies, `mode_is_trivial()` flags cases where the mode is not meaningful. It returns `TRUE` if all values are equally frequent. Modality is trivial in this case because it is a property of all values taken together, not of some values over others.

```{r}
mode_is_trivial(c("a", "b", "c"))
mode_is_trivial(c(1, 1, 2, 2, 3, 3))
mode_is_trivial(c(1, 1, 1, 2, 3))
```
