---
title: "Modal counts and frequencies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modal counts and frequencies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include=FALSE}
# Dev only: load moder from within moder
devtools::load_all(".")
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(moder)
```

In addition to finding a vector's modes, you might be interested in some metadata about them:

-   A vector's modal **count** is the number of its modes.

-   A vector's modal **frequency** is the number of times that any single mode appears in the vector.

This vignette lays out all the functions for modal metadata. In the end, it talks about a special feature of these functions, the `max_unique` argument.

## Modal count

`mode_count()` computes the number of modes:

```{r}
mode_count(c(5, 5, 6))
mode_count(c(5, 5, 6, 6, 7))
```

Even with missing values, the number of modes is sometimes known. It can only be 1 here because even if the `NA` is secretly `"b"`, then `"b"` would appear twice, but `"a"` would appear three times:

```{r}
mode_count(c("a", "a", "a", "b", NA))
```

All of this only works if the full set of modes can be determined. Below, `NA` could secretly be `7`, `8`, or any other value. If it's `8`, both numbers are equally frequent. Otherwise, `7` is the only mode. Since we lack this information, the number of modes is unknown.

```{r}
mode_count(c(7, 7, 7, 8, 8, NA))
```

Use `mode_count_range()` in such cases. It will determine the minimal and maximal number of modes, never returning `NA`. For more on `mode_count_range()`, see below, section *Maximal number of unique values*.

```{r}
mode_count_range(c(7, 7, 7, 8, 8, NA))
```

## Modal frequency

`mode_frequency()` counts the instances of a vector's modes in the vector:

```{r}
mode_frequency(c(4, 4, 5))
mode_frequency(c(4, 4, 4, 5))
```

Missing values are an issue here, even if the mode is obvious. Each `NA` might be another instance of the mode, so the frequency is unknown:

```{r}
mode_frequency(c(1, 1, 1, 1, 2, NA, NA))
```

With `mode_frequency_range()`, at least the minimal and maximal frequencies can be determined. It never returns `NA`. The minimum frequency supposes that no `NA`s represent the mode; the maximum frequency supposes that all of them do. In this way, there are four instances of `1` without counting the `NA`s, and six with counting them:

```{r}
mode_frequency_range(c(1, 1, 1, 1, 2, NA, NA))
```

### Uniform distribution

`is_uniform()` flags cases where all values are equally frequent. Modality is trivial in this case because it is a property of every single value, not of some values over others.

```{r}
is_uniform(c("a", "b", "c"))
is_uniform(c(1, 1, 2, 2, 3, 3))
is_uniform(c(1, 1, 1, 2, 3))
```

The mode is clearly not a useful concept in the first two cases [cf. @härdle2015, p. 40]. Some authors say that the mode is not defined if each value appears only once [@manikandan2011, p. 214]. However, it is certainly possible for the maximal frequency to be 1, so the only way for such distributions not to have any modal values would be a specific exception in the definition of the mode. The same applies to uniformly distributed data in general. No such exception appears in any definition that I am aware of. Even if it were to be suggested, I think the more elegant solution would be to accept all values of uniformly distributed data as trivially modal.

## Maximal number of unique values

All of moder's functions for metadata, such as `is_uniform()` and `mode_count_range()`, have a `max_unique` argument. It allows you to state how many unique values your data can have at the maximum. Why is this important? The two functions care about possible modes beyond the known values. In other words, their results might depend on whether or not the `NA`s can mask modal values that don't even occur among the known values! If that is possible, it presents an additional source of uncertainty.

Conversely, `max_unique` limits the possible number of such wildcard modes. Specify it as an integer that is the maximal number of unique values. If there can be no other values than those already known, specify `max_unique` as `"known"` instead. Always use `"known"` if you have factor data or you will get a warning. (The idea behind factors is that all possible values are known at the outset.)

Note that this argument does not represent an analytical decision but simply conveys your knowledge of the data to the computer. There is no meaningful choice to make: If the maximum number of unique values is known, you must specify `max_unique`; if not, you must not do so. Otherwise, you risk incorrect results if any values are missing. The default is `NULL` because the baseline assumption is always that nothing is known about missing values except for their number.

Below is an example. If two of the `NA`s represent `8` and the other three stand for a third value, all values appear with the same frequency, as determined by `is_uniform()`. This scenario is not certain at all, but it can't be ruled out either, so the function returns `NA`. As `mode_count_range()` shows, there could be three modes at most. (The minimum is always one if any values are missing.)

```{r}
x1 <- c(7, 7, 7, 8, NA, NA, NA, NA, NA)
is_uniform(x1)
mode_count_range(x1)
```

The picture is different if we know that each missing value must represent a known value, i.e., `7` or `8`. Even if two `NA`s stand for `8`, the other three can't be evenly distributed across `7` and `8`, so one of these values must be more frequent than the other one. This makes the distribution non-uniform. Also, there can only be one mode, so both the minimal and maximal mode counts are `1`.

```{r}
x1
is_uniform(x1, max_unique = "known")
mode_count_range(x1, max_unique = "known")
```

Three more functions have a `max_unique` parameter: `mode_count()`, `mode_frequency()`, and `mode_frequency_range()`. However, this only matters for corner cases. See [this Github issue](https://github.com/lhdjung/moder/issues/1).

# References
