---
title: "Performance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(moder)
```

Accuracy and flexibility are among the strong suits of moder. Performance is not. `mode_all()` and friends are slower than R's built-in functions for common measures of central tendency, `mean()` and `median()`. In the example below, the latter two take around 5 microseconds at the median on my MacBook Pro, while the mode functions take 35 to 60.

Why is that? All of the code is written in R. By contrast, the default `mean()` method calls an internal function. Also, handling missing values is much more complex when determining the mode(s) than with the mean or the median; see `vignette("missings")`. The default `mean()` and `median()` methods are fairly short when compared to `mode_first()`, and to `mode_all()` if its helper function is included. `mode_single()` is a wrapper around `mode_all()`, so it's even longer.

It is also possible that some of the code is less efficient than it could be. In particular, rewriting moder's R code in a language like C++ or Rust might blend accuracy with high performance. I myself am not going to port the package in the near term, so help is welcome. If you have any suggestions, please [open an issue](https://github.com/lhdjung/moder/issues) or get in touch via email (jung-lukas\@gmx.net).
