---
title: "Performance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(moder)
```

Accuracy and flexibility are among the strong suits of moder. Performance is not. `mode_single()` and friends are slower than R's built-in functions for common measures of central tendency, `mean()` and `median()`. In the example below, the latter two take around 5 microseconds at the median, while the mode functions take 35 to 60.

Why is that? All of the code is written in R. By contrast, the default `mean()` method calls an internal function. Also, handling missing values is much more complex when determining the mode(s) than with the mean or the median (see `vignette("missings")`). The default `mean()` and `median()` methods are fairly short when compared to `mode_first()`, and to `mode_all()` if its helper function is included. `mode_single()` is a wrapper around `mode_all()`.

It is also possible that some of the code is less efficient than it could be. If you have ideas for improvements, please [open an issue](https://github.com/lhdjung/moder/issues) or get in touch via email (jung-lukas\@gmx.net).

## Benchmarking

The mode functions returns `1` for the vector below, while `mean()` and `median()` return `NA`. I still think it makes sense to compare them because this difference may occur in many realistic distributions: Some values are missing, so the mean and the median are unknown, but there are enough known values to determine at least one mode.

```{r, eval=FALSE, include=FALSE}
# NOTE: The code below doesn't run when
# knitting the page, and I inserted
# the image manually. This is to avoid
# repetition of time-consuming benchmarking.
```

```{r, eval=FALSE}
library(bench)
library(ggplot2)

x <- c(2, 4, 6, NA, 3, 1, 1, 8, 1, 7)

df <- mark(
  mode_single = mode_single(x),
  mode_first = mode_first(x),
  mode_all = mode_all(x),
  mean = mean(x),
  median = median(x),
  check = FALSE, iterations = 100000
)

plot(df) +
  theme_minimal(base_size = 13)
```

![](images/moder_benchmarks)
